---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

\begin{center}
\vspace*{10\baselineskip}
\LARGE{\textbf{Project 1: PCA Analysis on a Birth Weight Dataset}}\\
\vspace*{19\baselineskip}
\normalsize{Trevor Carpenter, Christina De Cesaris, Michelle Tran}\\
STA 135: Multivariate Data Analysis\\
Professor Maxime Pouokam\\
University of California, Davis\\
May 4, 2021
\end{center}

\clearpage

```{r}
smok = read.csv("data/Birthweight_reduced_kg_R.csv")
```

# I. Introduction
Although low birth weighted infants can be healthy, under improper care or other circumstances, it can cause serious long-term health problems. A low birth weight is defined as being below 2.5kg. It can be caused by a multitude of factors \textemdash poor socioeconomic situation, premature birth, a mother's pre-existing health conditions, and more. In this project, we will highlight how smoking and other factors can cause low birth weights in infants in comparison to non-smoking parents. \
\
[state question of interest]. In order to answer this, we will be using principal component analysis to analyze the data. PCA is a statistical technique that reduces a dimensionality of a dataset to make it interpretable while also preserving the variability of the data. [more explanation?] We can use it to see what variables group together, identify possible outliers, recognize correlations between variables, and more. 




# II. Summary of Data
Before conducting PCA onto a dataset, we must first examine the data and decide on what variables we want to specifically look at by looking at different plots and summaries.  \

```{r}
#Plots




```
The plot above shows.... \
\
When conducting the analysis, we also want to know if we want to use the variance matrix or the correlation matrix. [Discuss the positive and negative aspects of using the covariance matrix for a PCA rather than the correlation matrix.]





# III. Analysis


1. 
[If the overall goal is to compare babies according with their mother’s smoking status, how would you proceed ?

My suggestion:  Run PCs on these two groups (use the smoking status to split the data and run PCs on them).Then comment on their similarities and differences.  You can compare their PCs, scree plots, etc ...]
```{r}
nonsmoke = subset(smok, smoker == 0)
smoke = subset(smok, smoker == 1)
#use: select = c("Length", "Birthweight", "other variables") in subset fcn for specific columns 
```



2.
[Determine the number of PCs needed when using the correlation matrix.]





3. 
[Interpret  the  PCs  chosen  from  the  previous  part. Make sure to specifically  comment  on  whether  positive  or negative scores (or scores close to 0) for a PC would likely be beneficial for baby.

Hint: To help see which of the loadings are “away from zero”, one could set the cutoff argument value in summary()to something other than 0.0.  For example one can use a cutoff of 0.2. ]




4.
[Examine plots of the PC scores and interpret them in the context of the problem. For example, what do you think of a particular patient (Baby).]




# IV. Interpretation
Rehash and summarize analysis section here.

[Suppose a new baby is born after the PCA has been completed. The newly born has the characteristics as indicated in Table 1.  Through using the previous PCA results, discuss how this particular individual/baby would compare to the other babies.]






# V. Conclusion
[What is your general conclusion and possible future work. By future work, I mean if you have more time what possible questions would you want to investigate ?]








\clearpage

\subsection*{Appendix: R Script}

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
