---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

\begin{center}
\vspace*{10\baselineskip}
\LARGE{\textbf{Project 1: PCA Analysis on a Birth Weight Dataset}}\\
\vspace*{19\baselineskip}
\normalsize{Trevor Carpenter, Christina De Cesaris, Michelle Tran}\\
STA 135: Multivariate Data Analysis\\
Professor Maxime Pouokam\\
University of California, Davis\\
May 4, 2021
\end{center}

\clearpage

```{r}
smok = read.csv("data/Birthweight_reduced_kg_R.csv")
#subset(smok, smoker == 1, select = c("Length", "Birthweight"))
```

# I. Introduction
[background info] \
In our project, we will highlight how smoking and other variables can lead to a decrease in birth weights in babies compared to non-smoking parents. [state question of interest]. In order to answer this, We will be using principal component analysis to analyze the data. \
PCA a statistical technique that reduces a dimensionality of a dataset to make it interpretable while also preserving the variability of the data. [more explanation?] 




# II. Summary of Data

```{r}
#Plots




```






# III. Analysis








# IV. Interpretation







# V. Conclusion










\clearpage

\subsection*{Appendix: R Script}

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
